Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-11-26 analysis/paper/paper.qmd
edited by the end of methods

tcsa_site <- tcsa %>% 
  group_by(site_name_top) %>% 
  add_tally() %>% 
  filter(n>1) %>% 
  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
  dplyr::select(min_tcsa,
         site_name_top, 
         label_site,
         n) %>% 
  distinct_all()

@@

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artifact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We distinguished the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the TCSA pattern. We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information. We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic people used quartzite and vein quartz for core and flake tools. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values for each raw material type with `r n_raw` artifacts to examine the interaction of raw material types and TCSA values. We categorized rare raw materials, which have less than 10 artifacts, as "Other".

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

### Temporal pattern

We used radiocarbon ages to explore the temporal pattern of the function of stemmed points. After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. Then we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other". We also explore possible environmental impact by connecting stemmed points on the eco-regional zones. We chose two maps of modern eco-regional zones and modified them for this research. @yim1975distribution first created vegetation forest map defined temperature and precipitations. This map then modified by @yi2010vegetation. They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW). We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, big river, island, and big mountain range in addition to ecological information including temperature and precipitation. We only focused on 14 zones, excluding two islands that have no evidence of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. While it is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, they are the only types of maps that are currently available to us.

@@

The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the techniques for making the new tools, and the new tools themselves can alter the typical contexts of tool use. For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip]. Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed. For example, @eren2022plains used TCSA to compare the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape. Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing. Folsom points show a narrower range of variation, indicating a they were more likely used for specific tasks.

Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, then people likely produced tools that performed a small set tasks. This may be due to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging]. We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.

>>>>

tcsa_site <- tcsa %>% 
  group_by(site_name_top) %>% 
  add_tally() %>% 
  filter(n>1) %>% 
  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
  dplyr::select(min_tcsa,
         site_name_top, 
         label_site,
         n) 
# %>% 
#   distinct_all()

@@

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artifact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We distinguished the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the TCSA pattern. We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information. We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Each analysis uses slightly different number of artifacts due to data availability.

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic, people used quartzite and vein quartz for core and flake tools. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values for each raw material type with `r n_raw` artifacts to examine the interaction of raw material types and TCSA values. We categorized rare raw materials, which have less than 10 artifacts, as "Other".

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to distinguish different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

### Temporal patterns

We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points. After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. Then we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other". We also explored possible environmental impact by connecting stemmed points to the eco-regional zones. We chose two maps of modern eco-regional zones and modified them for this research. @yim1975distribution first created vegetation forest map defined temperature and precipitations. This map then modified by @yi2010vegetation. They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW). We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation. We only focused on 14 zones, excluding two islands that have no records of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. It is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, but they are the only maps of eco-regional seperation that are currently available to us.

@@

The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the manufacturing techniques for the new tools. Moreover, the new tools themselves can alter the typical contexts of tool use. For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip]. Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed. For example, @eren2022plains used TCSA to compare the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape. Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing. Folsom points, on the other hand, show a narrower range of variation, indicating a they were more likely used for specific tasks.

Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the temporal or/and spatial patterns of TCSA range. If stemmed points represent a narrow TCSA range, then people likely produced tools that performed specific tasks. This may be related to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging]. We expect to observe various TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.
